\myTitle

### 2025-03-27 tex/responses-to-referees/responses-to-reviewers-1.Rmd

---
title: "EVA-2025-003-RA Cover Letter and Responses to Reviewers"
output: pdf_document
date: "2025-03-26"
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Thank you for the opportunity to revise our manuscript for _Evolutionary Applications_.
We have made a number of changes as recommended by the referees and feel that
these have greatly improved the manuscript.

In addition to the responses to referees (below) we are uploading a version of
the manuscript created with `latexdiff` that shows the changes we have made in
the document since the previously submitted version.

It has been a pleasure working with the journal's editorial staff.

@@

- Line 589: “unprecedented” is an over-statement as there is no independent validation set. The authors should use a word that less forceful.
    - _Changed to "excellent"_

- Sentence at lines 645-647. Can the authors provide a reference to back up this statement.
    - _We added a citation here that extensively describes the use of a Fluidigm chip-based panel for
    parent pair allocation in Chinook salmon in California._

- The online data and scripts (thank you for providing these) needs a proper readme file describing what the folders and files relate to.

    - A comprehensive README has been added.  The repo and its README can be viewed at:
    [https://anonymous.4open.science/r/california-chinook-microhaps-A661/README.md](https://anonymous.4open.science/r/california-chinook-microhaps-A661/README.md).  This will be archived at Zenodo upon acceptance.

### 2025-03-31 tex/responses-to-referees/responses-to-reviewers-1.Rmd

_Thank you for your comments and suggestions.  We have addressed all of them, as
detailed below, and we are pleased with how this has improved the manuscript._

### Major Comments

* I’m confused by the amount of sequencing and a clear purpose of each sequencing run that should be clarified. There was ddRAD to identify novel polymorphisms/microhaplotypes, the second run using a using PCR-GBS type sequencing has an unclear purpose. Was this on the full population? Another “amplicon sequencing” run was performed to convert SNPtype polymorphisms, which sounds similar to the second sequencing run of ddRAD polymorphisms but uses different terminology. Was this part of the previous run? Was this run a few samples or the full population? I presume these were different runs but maybe they were the same. The GT-seq method, does this include all samples and the full set of markers to this point?
    - _We added to this section to make it more clear that the ddRAD one was to find gene regions, and the
    subsequent sequencing runs were made on amplicons from those gene regions, and we are also explicit about
    how many fish were used in those GT-seq optimization runs._

* Authors provide considerable detail on sample location in the methods but only provide the sample number in the introduction. I believe it would be useful to have summary sentences of Table 1 that include number sites for spring, fall, etc run, and the number of samples within each ecotype, total samples within the methods, etc.
    - _Thank you for this suggestion.  We added the following to the end of the first paragraph in the Population Sampling portion of the
Methods section:_
"Of the 1,636 total fish samples from 17 different populations, 727 fish from 10 collections were of the fall-run ecotype,
498 fish from 5 collections were spring run, and 111 and 300 fish were of the winter-run and late-fall-run ecotypes, 
respectively,  each represented by a single collection."

### 2025-03-31 tex/abstract.tex
Updates/edits from Carlos

Genetic methods have become an essential component of ecological investigation
and conservation planning for fish and wildlife. Among these methods is the use
of genetic marker data to identify individuals to
populations, or stocks, of origin. More recently, methods that involve genetic
pedigree reconstruction to identify relationships between individuals within populations
have also become common. We present, here, a novel set of
multi-allelic microhaplotype genetic markers for Chinook salmon which provide unprecedented
resolution for population discrimination and relationship identification from a rapidly and
economically assayed panel of markers. We show how this set of microhaplotypes
provides definitive power to identify all known lineages of Chinook salmon in
California. The inclusion of genetic loci that have known associations with phenotype
and that were identified as outliers in examination of whole genome sequence data,
allows resolution of stocks that are not highly genetically differentiated but
are phenotypically distinct and managed as such. 
This same set of multiallelic genetic markers have ample variation to
accurately identify parent-offspring and full-sibling
pairs in all California populations, including the genetically depauperate winter-run
lineage.
Validation of this marker panel in coastal salmon populations not previously studied with
modern genetic methods, also reveals novel biological insights, including
the presence of a single copy of a haplotype for a phenotype that has not
been documented in that part of the species range, and a clear signal of mixed
ancestry for a salmon population that is on the geographic margins of the 
primary evolutionary lineages present in California.

>>>>

Genetic methods have become an essential component of ecological investigation
and conservation planning for fish and wildlife. Among these methods is the use
of genetic marker data to identify individuals to
populations, or stocks, of origin. More recently, methods that involve genetic
pedigree reconstruction to identify relationships between individuals within populations
have also become common. We present, here, a novel set of
multi-allelic microhaplotype genetic markers for Chinook salmon which provide excellent
resolution for population discrimination and relationship identification from a rapidly and
economically assayed panel of markers. We show how this set of genetic markers assayed
by sequencing of 204 amplicons, in tandem with a reference data set of 1,636 individual
samples from 17 populations,
provides definitive power to identify all known lineages of Chinook salmon in
California. The inclusion of genetic loci that have known associations with phenotype
and that were identified as outliers in examination of whole genome sequence data,
allows resolution of stocks that are not highly genetically differentiated but
are phenotypically distinct and managed as such.
This same set of multiallelic genetic markers have ample variation to
accurately identify parent-offspring and full-sibling
pairs in all California populations, including the genetically depauperate winter-run
lineage.
Validation of this marker panel in coastal salmon populations not previously studied with
modern genetic methods, also reveals novel biological insights, including
the presence of a single copy of a haplotype for a phenotype that has not
been documented in that part of the species range, and a clear signal of mixed
ancestry for a salmon population that is on the geographic margins of the
primary evolutionary lineages present in California.

### 2025-03-31 tex/body-text-of-paper.tex
Updates/edits from Carlos

Samples of fish for the reference baseline were compiled from 13 locations within California, and one
within Oregon. This set of populations and stocks includes all of the previously described lineages of Chinook salmon in the southern portion of the range and most of the populations that are known to have significant genetic differentiation in this region.
These locations included sites in the two
main tributaries (the Sacramento River and the San Joaquin River) within
California's Central Valley, and the two main tributaries (Klamath and Trinity Rivers) in the Klamath basin in Northern California, and
three rivers of the California coast (Russian, Eel, and Smith).
Of the 1636 total fish samples from 17 different populations, 727 fish from 10 collections were of the fall run ecotype,
498 fish from 5 collections were spring run, and 111 and 300 fish were of the winter run and late-fall run ecotypes, 
respectively,  each represented by a single collection.
(Figure~\ref{fig:map}, Table~\ref{tab:samples}).
%%%%%%%%%%%%%
\begin{figure}
\newcommand{\mapcap}{\footnotesize  A map of sampling locations represented in the reference
baseline.  The colored circles aside
each location name represent the different run-timing groups of Chinook salmon
included in the collections from the location.  Color and shape codes are as follows: \Wball~=~Winter run,
\LFball{}~=~Late-fall run, \Fball~=~Fall run, \Sball~=~Spring run.
Names of each location are followed by the location codes (Table~\ref{tab:samples}).
Abbreviations used in location names as follows:
R. = River; Ck. = Creek; H. = Hatchery; NFH = National Fish Hatchery.}
\begin{center}
\includegraphics[width=0.48\textwidth]{images/map-crop.pdf}
\end{center}
\caption[\mapcap]{\mapcap}
\label{fig:map}
\end{figure}
%%%%%%%%%%%%%%%

Collections in each location were separated into fish from the different adult migration timing ecotypes according
to a variety of criteria.  Winter-run Chinook salmon are propagated at the Livingston
Stone National Fish Hatchery for supplementation (released as juveniles into the river) and as a captive broodstock;
sampling of this population was performed by hatchery staff during spawning. In Central Valley rivers with both
spring-run and fall-run ecotypes (i.e. Mill, Deer and Butte Creeks), previous research has documented different  (but overlapping) ranges of spawning time (Julian date) for each ecotype \citep{fry1961king,yoshiyama1998historical}.
As such, sampling targeted carcasses found in each of these
rivers at times representative of their run-timing designation, which were further confirmed using their genotypes
within the region of strongest association (RoSA) on chromosome 28 \citep{thompson2020complex}.
In the San Joaquin River, in 2016, at the time of sampling, only
fall-run Chinook salmon adults were found, therefore the samples could be categorized.  Run-timing
designation at the Feather River Hatchery is more complicated but it is managed by the hatchery practices in place:
early returning fish in May and April (expressing the spring-run life history) are visibly tagged and returned
to the river. Subsequently in the fall, when hatchery raceways are reopened, only fish with the visible tag are included
as spring-run broodstock, and the rest are included in the fall-run broodstock.
Putative spring-run and fall-run fish 
from these two broodstocks were sampled. Similar to broodstock collections in the CCV, samples from spring- and fall-run fish in the Trinity River were selected on the basis of spawn timing at the Trinity River Hatchery with migration-timing ecotype confirmed by RoSA genotype. Elsewhere, adult samples were taken from live broodstock at Rowdy Creek (Smith River), Iron Gate (Klamath River) and Warm Springs (Russian River) Hatcheries and juveniles were sampled from Blue Creek (Klamath River); all locations where only fall-run fish have been documented.  Spring run adult fish were sampled from the spring-run broodstock at Cole Rivers Hatchery on the Rogue River. Samples from the Eel River were taken from live fish ascending a ladder in the upper basin and were confirmed as fall-run ecotype by RoSA genotype.

@@

The 8,725 discovered RAD loci were filtered down to a set of markers suitable for conversion
to amplicon sequencing assays. Gene regions were selected from the RAD loci that (1) contained multiple SNPs within a 100 bp window,
and (2) showed haplotype variation amongst the winter run and at least one other population.
A 100 bp window was used because the success of multiplex PCR and MiSeq sequencing is
diminished in amplicons larger than 100 bp. Primer3
software \citep{untergasser2012primer3} was used to design primers for 229 gene regions that met these criteria.
These 229 gene regions were tested for PCR-amplification and consistency in a separate sample
of 288 Chinook salmon: 96 SRW fish and 48 each from the Entiat River (a Columbia River
tributary), TRH, FRHS, and CRLF\@.  Eight GT-seq amplifications were made and these were sequenced
on eight runs of a MiSeq sequencer
using a $2\times 75$-cycle kit to identify problematic loci and to optimize primer dilutions.
The sequences from each run at each locus were mapped to the {\em de novo} reference sequences (produced by Stacks) with
the Burrows-Wheeler (bwa) aligner using the mem algorithm \citep{bwa-mem2009}. 
Subsequently, alignments were sorted with SAMtools \citep{li2009sequence} and provided to
FreeBayes \citep{garrison2012haplotype} for variant calling.
FreeBayes output was filtered to remove indels and retain variants with a Phred-scaled quality score of 30
or greater and a read depth of 10 or higher. Further filtering and analysis of the loci, including
assessment of Hardy-Weinberg equilibrium \citep{hardy1908mendelian} was conducted in MICROHAPLOT (Ng {\em et al.}, DOI: 10.5281/zenodo.820110).  Loci that produced more than two haplotypes within an individual, and those clearly violating Hardy-Weinberg equilibrium within populations were removed.

@@

A number of SNP markers from an earlier discovery effort \citep{clemento2011discovery} had already
been validated as useful for GSI
and PBT in California \citep{clemento2014evaluation}. These markers were originally assayed
using TaqMan (Life Technologies Corp., Carlsbad, CA) probes or SNPtype assays (Standard BioTools Inc., South San Francisco, CA).
These markers were typed using amplicon sequencing, following the methods in
\citet{campbell2015genotyping} with the addition of the Illumina small RNA sequencing primer and
read 2 sequencing primer to the SNPtype Specific Target Amplification Primer and Locus
Specific Primers respectively.  Loci were removed that were frequently inconsistent with SNPtype assays,
had poor amplification, or had excessive amplification and were subsequently over-represented in the sequencing library.  FreeBayes was used to identify all SNPs (additional to the TaqMan-assay targets) along the fragment, which facilitated scoring fragment microhaplotypes when multiple SNPs were present along the read \citep{baetscher2018microhaplotypes}. The resulting genetic markers are hereafter referred to as `SNPlicons' to record their origin as TaqMan assays in a previous
baseline.

@@

Initial examination of sequence data 
in the 16 winter-run fish and 84 non-winter run fish revealed
a number of SNPs that were potentially diagnostic, but none were fixed for alternate alleles 
in a larger sample.
Nonetheless, three of these markers had large frequency differences
between the groups and are included in the reference baseline.  The process of 
discovering them is described in the supplemental text section \ref{sec:wrap-methods}.

@@

At values of $K$ from 2 to 7, clusters identified by the program STRUCTURE
generally corresponded to groupings of related populations, and confirmed
{\em a priori} knowledge about which collections in the baseline could and should be
grouped together into reporting units (Figure~\ref{fig:struct}).  At every value of $K$, CLUMPAK discerned at least
12 of 20 replicates in the major mode (Figure~\ref{fig:struct}).   
Clustering solutions in the minor
modes generally converged on a single alternative solution and appear in Figure~\ref{fig:minor-modes}.
%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\newcommand{\structcap}{\footnotesize Major modes found by STRUCTURE and summarized
by CLUMPAK out of 20 replicates for each value of $K$, the assumed number of
subpopulations. $K$ value is listed at lower right of each barplot, followed by the number
of replicate runs of structure that converged to this major mode.}
\begin{center}
\includegraphics[width=0.48\textwidth]{images/clumpak-crop.pdf}
\end{center}
\caption[\structcap]{\structcap}
\label{fig:struct}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%
The late-fall run collection (CHLF) appears as a separate cluster at $K$ values
as low as 4, with the inclusion of the LFAR markers.  This is notable, since accurately discriminating between the late-fall and
fall-run Chinook salmon of the Central Valley has been impossible with previously used genetic marker sets.
Another noteworthy feature appears at $K=7$, with STRUCTURE separating the fall-run reporting unit
into two clusters that do not correspond exactly with the individual
collections within the reference baseline. For example, fish from the
cluster that predominates in the FRHS collection are also found in the FRHF and MDS collections.

@@

Overall, of the 1636 fish in the reference baseline, 1612 (98.5\%) were correctly
assigned to their reporting unit of origin, while 24 (1.5\%) were incorrectly assigned.
Four of those 24 misassignments which involved fish being incorrectly assigned to
SRW, were clearly fish from SRW that had been incorrectly sampled into a different
collection, likely due to straying. (This is evident because SRW are so distinct that it is highly improbable they
would be incorrectly identified).  Of the remaining 20 misassignments,  8 involved fish from the
CHLF (late fall run) collection being misassigned to the CV-Fall reporting unit and 6 were
fish from the MDS (Mill-Deer spring run) collection being assigned to the CV-Fall reporting unit.
It is possible that some of these misassignments represent CV-Fall fish that were incorrectly
sampled as CHLF or MDS; however, unlike with SRW,  we  cannot conclude that
confidently, as there is some degree of overlap in likelihood values between the two.

@@

The presence of multiple alleles at many of these loci improves power for
kin inference compared to a marker panel using only a single SNP at each
amplicon.  To provide a visual display of the difference afforded by using
microhaplotypes compared to single SNPs in this data set, we show the
distribution of log-likelihood ratios for different relationships using our microhaplotype-scored
amplicons, versus using the most heterozygous single SNP from each
amplicon for all fish from the FRH (Fig.~\ref{fig:ckmr-comp}), with results for the PO vs Unrelated
log likelihood ratio shown for more collections in Fig.~\ref{fig:ckmr-comp-supp}.  The figures show a considerable increase
in separation between the relationship categories from calling genotypes
as microhaplotypes in these amplicons.
%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
\newcommand{\fprcap}{\footnotesize The distribution of log-likelihood ratios for
fish from Feather River Hatchery when scoring markers as microhaplotypes versus as
single SNPs.  Three different likelihood ratios are shown in the three panels.  The opacity
of the distributions indicates the marker type: the more opaque setting indicates markers
scored as microhaplotypes; the more transparent setting indicates the use of only the most
heterozygous SNP from amongst all SNPs in the microhaplotype.  Evidently, scoring the markers
as microhaplotypes produces more separation between the distributions for the relationships of PO, FS, or HS, against
the relationship of unrelated (U).}
\begin{center}
\includegraphics[width=0.5\textwidth]{images/frh-comp-figure-crop.pdf}
\end{center}
\caption[\fprcap]{\fprcap}
\label{fig:ckmr-comp}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%

@@

We describe a novel panel of microhaplotype genetic markers for Chinook salmon
in California that provides sufficient power for highly accurate identification 
of parent and offspring pairs and full siblings, and also allows near-perfect
identification of individuals to population or genetic group of origin in California. 
This set of genetic markers includes multiallelic gene regions with high variability for
relationship inference, gene regions identified in whole-genome sequence data
for increased power for identification of specific populations, and markers that have
long been in use for both GSI and PBT in California \citep{clemento2014evaluation}, by converting them
into so-called 'SNPlicons,' which are treated as multiallelic haplotypes in the presence
of additional variation surrounding the originally assayed SNP\@. Our work adds to a growing
chorus of recent studies finding that microhaplotype markers can provide additional
power over single SNPs for both GSI \citep{mckinney2017managing,mckinney2022new,hargrove2024efficient} and
relationship inference \citep{baetscher2018microhaplotypes,bootsma2020gt,delomas2024microhaplotypes}.
This set of markers lays the foundation for a comprehensive 
genetic monitoring and evaluation effort that facilitates multiple types of inference 
and is flexible and extensible.

@@

Moreover, we show how the FRH spring-run lineage is
easily identifiable through a combination of traditional GSI and the characterization
of functional genetic markers in the RoSA. Finally, although previous work has described
GSI capabilities that distinguish the natural-spawning CCV spring-run lineages from
each other and their fall-run counterparts with moderately high accuracy, we demonstrate
near complete accuracy in distinguishing these 'stocks'. Moreover, the few fish that are
apparently misidentified (Figure~\ref{fig:gsi}) are likely to be primarily migrants and not true misidentifications.
For example, the three fish that were field-characterized as spring-run from Butte Creek, but are
genetically identified as winter-run salmon clearly carry the winter-run genomic background and
are strays from the winter-run stock, as these could not realistically be misidentified on the 
basis of the genetic data.

>>>>

Samples of fish for the reference baseline were compiled from 13 locations within California, and one
within Oregon. This set of populations and stocks includes all of the previously described lineages of Chinook salmon in the southern portion of the range and most of the populations that are known to have significant genetic differentiation in this region.
These locations included sites in the two
main tributaries (the Sacramento River and the San Joaquin River) within
California's Central Valley, and the two main tributaries (Klamath and Trinity Rivers) in the Klamath basin in Northern California, and
three rivers of the California coast (Russian, Eel, and Smith).
Of the 1,636 total fish samples from 17 different populations, 727 fish from 10 collections were of the fall-run ecotype,
498 fish from 5 collections were spring run, and 111 and 300 fish were of the winter-run and late-fall-run ecotypes, 
respectively,  each represented by a single collection.
(Figure~\ref{fig:map}, Table~\ref{tab:samples}).
%%%%%%%%%%%%%
\begin{figure}
\newcommand{\mapcap}{\footnotesize  A map of sampling locations represented in the reference
baseline.  The colored circles aside
each location name represent the different run-timing groups of Chinook salmon
included in the collections from the location.  Color and shape codes are as follows: \Wball~=~Winter run,
\LFball{}~=~Late-fall run, \Fball~=~Fall run, \Sball~=~Spring run.
Names of each location are followed by the location codes (Table~\ref{tab:samples}).
Abbreviations used in location names as follows:
R. = River; Ck. = Creek; H. = Hatchery; NFH = National Fish Hatchery.}
\begin{center}
\includegraphics[width=0.48\textwidth]{images/map-crop.pdf}
\end{center}
\caption[\mapcap]{\mapcap}
\label{fig:map}
\end{figure}
%%%%%%%%%%%%%%%

Collections in each location were separated into fish from the different adult migration timing ecotypes according
to a variety of criteria.  Winter-run Chinook salmon are propagated at the Livingston
Stone National Fish Hatchery for supplementation (released as juveniles into the river) and as a captive broodstock;
sampling of this population was performed by hatchery staff during spawning. In Central Valley rivers with both
spring-run and fall-run ecotypes (i.e. Mill, Deer and Butte Creeks), previous research has documented different  (but overlapping) ranges of spawning time (Julian date) for each ecotype \citep{fry1961king,yoshiyama1998historical}.
As such, sampling targeted carcasses found in each of these
rivers at times representative of their run-timing designation, which were further confirmed using their genotypes
within the region of strongest association (RoSA) on chromosome 28 \citep{thompson2020complex}.
In the San Joaquin River, in 2016, at the time of sampling, only
fall-run Chinook salmon adults were found, therefore the samples could be categorized.  Run-timing
designation at the Feather River Hatchery is more complicated but it is managed by the hatchery practices in place:
early returning fish in May and April (expressing the spring-run life history) are visibly tagged and returned
to the river. Subsequently in the fall, when hatchery raceways are reopened, only fish with the visible tag are included
as spring-run broodstock, and the rest are included in the fall-run broodstock.
Putative spring-run and fall-run fish 
from these two broodstocks were sampled. Similar to broodstock collections in the CCV, samples from spring- and fall-run fish in the Trinity River were selected on the basis of spawn timing at the Trinity River Hatchery with migration-timing ecotype confirmed by RoSA genotype. Elsewhere, adult samples were taken from live broodstock at Rowdy Creek (Smith River), Iron Gate (Klamath River) and Warm Springs (Russian River) Hatcheries and juveniles were sampled from Blue Creek (Klamath River); all locations where only fall-run fish have been documented.  Spring-run adult fish were sampled from the spring-run broodstock at Cole Rivers Hatchery on the Rogue River. Samples from the Eel River were taken from live fish ascending a ladder in the upper basin and were confirmed as fall-run ecotype by RoSA genotype.

@@

The 8,725 discovered RAD loci were filtered down to a set of markers suitable for conversion
to amplicon sequencing assays. Gene regions were selected from the RAD loci that (1) contained multiple SNPs within a 100 bp window,
and (2) showed haplotype variation amongst the winter run and at least one other population.
A 100 bp window was used because the success of multiplex PCR and MiSeq sequencing is
diminished in amplicons larger than 100 bp. Primer3
software \citep{untergasser2012primer3} was used to design primers for 229 gene regions that met these criteria.
These 229 gene regions were tested for PCR-amplification and consistency in a separate sample
of 288 Chinook salmon: 96 SRW fish and 48 each from the Entiat River (a Columbia River
tributary), TRH, FRHS, and CRLF\@.  Eight GT-seq amplifications were made and these were sequenced
on eight runs of a MiSeq sequencer
using a $2\times 75$-cycle kit to identify problematic loci and to optimize primer dilutions.
The sequences from each run at each locus were mapped to the {\em de novo} reference sequences (produced by Stacks) with
bwa mem \citep{bwa-mem2009}. 
Subsequently, alignments were sorted with SAMtools \citep{li2009sequence} and provided to
FreeBayes \citep{garrison2012haplotype} for variant calling.
FreeBayes output was filtered to remove indels and retain variants with a Phred-scaled quality score of 30
or greater and a read depth of 10 or higher. Further filtering and analysis of the loci, including
assessment of Hardy-Weinberg equilibrium \citep{hardy1908mendelian} was conducted in MICROHAPLOT (Ng {\em et al.}, DOI: 10.5281/zenodo.820110).  Loci that produced more than two haplotypes within an individual, and those clearly violating Hardy-Weinberg equilibrium within populations were removed.

@@

A number of SNP markers from an earlier discovery effort \citep{clemento2011discovery} had already
been validated as useful for GSI
and PBT in California \citep{clemento2014evaluation}. These markers were originally assayed
using TaqMan (Life Technologies Corp.) probes or SNPtype assays (Standard BioTools Inc.).
These markers were typed using amplicon sequencing, following the methods in
\citet{campbell2015genotyping} with the addition of the Illumina small RNA sequencing primer and
read 2 sequencing primer to the SNPtype Specific Target Amplification Primer and Locus
Specific Primers respectively.  Loci were removed that were frequently inconsistent with SNPtype assays,
had poor amplification, or had excessive amplification and were subsequently over-represented in the sequencing library.  FreeBayes was used to identify all SNPs (additional to the TaqMan-assay targets) along the fragment, which facilitated scoring fragment microhaplotypes when multiple SNPs were present along the read \citep{baetscher2018microhaplotypes}. The resulting genetic markers are hereafter referred to as `SNPlicons' to record their origin as TaqMan assays in a previous
baseline.

@@

Initial examination of sequence data 
in the 16 winter-run fish and 84 non-winter-run fish revealed
a number of SNPs that were potentially diagnostic, but none were fixed for alternate alleles 
in a larger sample.
Nonetheless, three of these markers had large frequency differences
between the groups and are included in the reference baseline.  The process of 
discovering them is described in the supplemental text section \ref{sec:wrap-methods}.

@@

At values of $K$ from 2 to 7, clusters identified by the program STRUCTURE
generally corresponded to groupings of related populations, and confirmed
{\em a priori} knowledge about which collections in the baseline could and should be
grouped together into reporting units (Figure~\ref{fig:struct}).  At every value of $K$, CLUMPAK discerned at least
12 of 20 replicates in the major mode (Figure~\ref{fig:struct}).   
Clustering solutions in the minor
modes generally converged on a single alternative solution and appear in Figure~\ref{fig:minor-modes}.
%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\newcommand{\structcap}{\footnotesize Major modes found by STRUCTURE and summarized
by CLUMPAK out of 20 replicates for each value of $K$, the assumed number of
subpopulations. $K$ value is listed at lower right of each barplot, followed by the number
of replicate runs of structure that converged to this major mode.}
\begin{center}
\includegraphics[width=0.48\textwidth]{images/clumpak-crop.pdf}
\end{center}
\caption[\structcap]{\structcap}
\label{fig:struct}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%
The late-fall-run collection (CHLF) appears as a separate cluster at $K$ values
as low as 4, with the inclusion of the LFAR markers.  This is notable, since accurately discriminating between the late-fall and
fall-run Chinook salmon of the Central Valley has been impossible with previously used genetic marker sets.
Another noteworthy feature appears at $K=7$, with STRUCTURE separating the fall-run reporting unit
into two clusters that do not correspond exactly with the individual
collections within the reference baseline. For example, fish from the
cluster that predominates in the FRHS collection are also found in the FRHF and MDS collections.

@@

Overall, of the 1,636 fish in the reference baseline, 1,612 (98.5\%) were correctly
assigned to their reporting unit of origin, while 24 (1.5\%) were incorrectly assigned.
Four of those 24 misassignments which involved fish being incorrectly assigned to
SRW, were clearly fish from SRW that had been incorrectly sampled into a different
collection, likely due to straying. (This is evident because SRW are so distinct that it is highly improbable they
would be incorrectly identified).  Of the remaining 20 misassignments,  8 involved fish from the
CHLF (late-fall run) collection being misassigned to the CV-Fall reporting unit and 6 were
fish from the MDS (Mill-Deer spring run) collection being assigned to the CV-Fall reporting unit.
It is possible that some of these misassignments represent CV-Fall fish that were incorrectly
sampled as CHLF or MDS; however, unlike with SRW,  we  cannot conclude that
confidently, as there is some degree of overlap in likelihood values between the two.

@@

The presence of multiple alleles at many of these loci improves power for
kin inference compared to a marker panel using only a single SNP at each
amplicon.  To provide a visual display of the difference afforded by using
microhaplotypes compared to single SNPs in this data set, we show the
distribution of log-likelihood ratios for different relationships using our microhaplotype-scored
amplicons, versus using the most heterozygous single SNP from each
amplicon for all fish from the FRH (Figure~\ref{fig:ckmr-comp}), with results for the PO vs Unrelated
log likelihood ratio shown for more collections in Figure~\ref{fig:ckmr-comp-supp}.  The figures show a considerable increase
in separation between the relationship categories from calling genotypes
as microhaplotypes in these amplicons.
%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
\newcommand{\fprcap}{\footnotesize The distribution of log-likelihood ratios for
fish from Feather River Hatchery when scoring markers as microhaplotypes versus as
single SNPs.  Three different likelihood ratios are shown in the three panels.  The opacity
of the distributions indicates the marker type: the more opaque setting indicates markers
scored as microhaplotypes; the more transparent setting indicates the use of only the most
heterozygous SNP from amongst all SNPs in the microhaplotype.  Evidently, scoring the markers
as microhaplotypes produces more separation between the distributions for the relationships of PO, FS, or HS, against
the relationship of unrelated (U).}
\begin{center}
\includegraphics[width=0.5\textwidth]{images/frh-comp-figure-crop.pdf}
\end{center}
\caption[\fprcap]{\fprcap}
\label{fig:ckmr-comp}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%

@@

We describe a novel panel of microhaplotype genetic markers for Chinook salmon
in California that provides sufficient power for highly accurate identification 
of parent and offspring pairs and full siblings, and also allows near-perfect
identification of individuals to population or genetic group of origin in California. 
This set of genetic markers includes multiallelic gene regions with high variability for
relationship inference, gene regions identified in whole-genome sequence data
for increased power for identification of specific populations, and markers that have
long been in use for both GSI and PBT in California \citep{clemento2014evaluation}, by converting them
into so-called 'SNPlicons,' which are treated as multiallelic haplotypes in the presence
of additional variation surrounding the originally assayed SNP\@. Our work adds to a growing
number of recent studies finding that microhaplotype markers can provide additional
power over single SNPs for both GSI \citep{mckinney2017managing,mckinney2022new,hargrove2024efficient} and
relationship inference \citep{baetscher2018microhaplotypes,bootsma2020gt,delomas2024microhaplotypes}.
This set of markers lays the foundation for a comprehensive 
genetic monitoring and evaluation effort that facilitates multiple types of inference 
and is flexible and extensible.

@@

Moreover, we show how the FRH spring-run lineage is
easily identifiable through a combination of traditional GSI and the characterization
of functional genetic markers in the RoSA. Finally, although previous work has described
GSI capabilities that distinguish the natural-spawning CCV spring-run lineages from
each other and their fall-run counterparts with moderately high accuracy, we demonstrate
near complete accuracy in distinguishing these 'stocks'. Moreover, the few fish that are
apparently misidentified (Figure~\ref{fig:gsi},~\ref{fig:eighty}) are likely to be primarily migrants and not true misidentifications.
For example, the three fish that were field-characterized as spring-run from Butte Creek, but are
genetically identified as winter-run salmon clearly carry the winter-run genomic background and
are strays from the winter-run stock, as these could not realistically be misidentified on the 
basis of the genetic data.
